Reasonably Programmable Literal Notation

### 2018-02-26 icfp18/syntax-icfp18.tex
new figure 3 -- regexes

\paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) is semantically expressive enough for integration into Reason/OCaml and other full-scale statically typed functional languages. We evaluate these claims with (1) a series of increasingly ML-like calculi that are equipped with proofs of these abstract reasoning principles; and (2) a number of non-trivial examples that appear throughout the paper and involve the language features mentioned above. In describing these examples, we also demonstrate that literal parsing logic can be defined using off-the-shelf parser generators, so the burden on notation providers is comparable to that of existing systems. % From the client's perspective, which is the focus of our contributions, TLMs are uniquely reasonable. 
Unique to our approach in this paper is the focus, both formal and informal, on maintaining client-side abstract reasoning principles.

@@

%\footnote{In general, spliced expressions might themselves apply TLMs, in which case the convention is to use a distinct color for unspliced segments at each depth. For example, if strings were not primitive, we might write \li{`SURL<body>\{EURLheading ($str "SSTRWorld!ESTR")SURL\} ...</body>EURL`}.}
\begin{figure*}[t]

\paragraph{Outline} 
% \begin{itemize}
% \item 
We begin in Sec.~\ref{sec:setsms} by introducing TLMs with a detailed case study. 
% \item 
Sec.~\ref{sec:sptsms} then introduces {pattern TLMs}, i.e. TLMs that generate patterns, and describes the special hygiene conditions in pattern position. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec.~\ref{sec:ptsms}  introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that can take type and module parameters, which allow us to define literal notation over a type- or module-parameterized family of types. Sec.~\ref{sec:ptsms} also reveals that the \li{expansions require} clause is syntactic sugar for parameterization followed immediately by partial parameter application. 
Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with additional examples that demonstrate both the expressive power of this approach and certain alternative parser implementation strategies that notation providers might consider. 
Sec.~\ref{sec:setsms-formally} presents a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above, in the formally simplest setting. 
Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives the most general variant of the reasoning principles theorem. 
% \item 
Sec.~\ref{sec:existing-approaches} compares TLMs to related work, guided by the rubric of five reasoning principles just discussed. 
% \item 
Sec.~\ref{sec:discussion} concludes with a summary of the contributions of this paper and a discussion of future work.

% Parameters serve two important purposes. First, they allow a TLM to operate not just at a single type, but over a type- and module-parameterized family of types. We give as an example a parametric TLM for string-keyed dictionaries that operates uniformly over different implementations of the abstract dictionary signature, and also over different choices of the range type. Second, parameters lower the syntactic cost of the strict context independence discipline that TLMs enforce because they give expansions access to terms other than those explicitly spliced in to the literal body by the client, and because %This is particularly convenient for working with abstract types \emph{a la} ML. 
% partial parameter application is possible. % Sec. \ref{sec:ptsms} culminates with a calculus that extends the calculus of Sec. \ref{sec:setsms-formally} with type functions, ML-style modules and parametric TLMs.
% \item
% \item 
% \end{itemize}

Due to the space limitations, certain technical details and extensions to the calculi presented in this paper, as well as detailed proofs, are available in the supplemental material. % Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators. %are also available in the supplement. Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e. \emph{static evaluation}, in  more detail. Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.

@@

\subsection{HTML}

>>>>

\paragraph{Contributions} This paper introduces \emph{typed literal macros} (TLMs): the first system for defining new literal notation that (1) provably maintains all five of the abstract reasoning principles just outlined; and (2) is semantically expressive enough for integration into Reason/OCaml and other full-scale statically typed functional languages. We evaluate these claims with (1) a series of increasingly ML-like calculi equipped with proofs of these abstract reasoning principles; and (2) a number of non-trivial examples that appear throughout the paper and involve the language features mentioned above. In describing these examples, we also demonstrate that literal parsing logic can be defined using off-the-shelf parser generators, so the burden on notation providers is comparable to that of existing systems. % From the client's perspective, which is the focus of our contributions, TLMs are uniquely reasonable. 
Unique to our approach in this paper is the focus, both formal and informal, on maintaining client-side abstract reasoning principles.

@@

%\footnote{In general, spliced expressions might themselves apply TLMs, in which case the convention is to use a distinct color for unspliced segments at each depth. For example, if strings were not primitive, we might write \li{`SURL<body>\{EURLheading ($str "SSTRWorld!ESTR")SURL\} ...</body>EURL`}.}

\begin{figure}[t]
\begin{subfigure}[t]{0.45\textwidth}
\begin{lstlisting}[mathescape=~]
  module Regex = {
    type t = AnyChar | Str(string)
           | Seq(t, t) | Or(t, t) 
           | Star(t);
  };
\end{lstlisting}
\vspace{-5px}
\caption{The \li{Regex} module, which defines the recursive datatype \li{Regex.t}.}
\label{fig:Regex-module-def}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.53\textwidth}
\begin{lstlisting}[mathescape=|]
  notation $regex at Regex.t
    lexer   RegexLexer
    parser  RegexParser.start
    expansions require
      module Regex as Regex;
\end{lstlisting}
\vspace{-5px}
\caption{The definition of the \li{$regex} TLM. Figure \ref{fig:candidate-exp-Reason} defines \li{RegexLexer} and \li{RegexParser}.}
\label{fig:html-tlm-def}
\end{subfigure}

\begin{subfigure}[t]{\textwidth}
\vspace{4px}
\begin{lstlisting}
  let any_base = $regex `(SURLA|T|G|CEURL)`;
  let bisA = $regex `(SURLGC$(EURLany_baseSURL)GCEURL)`;
  let gene_restriction_template(gene_sequence : string) => 
    $regex `(SURL$(EURLbisASURL)$(EURLany_baseSURL)*$$(EURLgene_sequenceSURL)$(EURLany_baseSURL)*$(EURLbisASURL)EURL)`
\end{lstlisting}
\vspace{-2px}
\caption{Examples of the \li{$regex} TLM being applied in a bioinformatics application. %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
}
\label{fig:first-tsm-example}
\end{subfigure}
\vspace{5px}
\caption{Case Study: POSIX-style regex literal notation, with support for string and regex splicing.}
\end{figure}

\paragraph{Outline} 
% \begin{itemize}
% \item 
We begin in Sec.~\ref{sec:setsms} by introducing TLMs with a detailed case study. 
% \item 
Sec.~\ref{sec:sptsms} then introduces {pattern TLMs}, i.e. TLMs that generate patterns, and describes the special hygiene conditions in pattern position. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec.~\ref{sec:ptsms}  introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that can take type and module parameters, which allow us to define literal notation over a type- or module-parameterized family of types. Sec.~\ref{sec:ptsms} also reveals that the \li{expansions require} clause is syntactic sugar for parameterization followed immediately by partial parameter application. 
Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with additional examples that demonstrate both the expressive power of this approach and alternative parser implementation strategies that notation providers can consider. 
Sec.~\ref{sec:setsms-formally} presents a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above, in the formally simplest setting. 
Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives the most general variant of the reasoning principles theorem. 
% \item 
Sec.~\ref{sec:existing-approaches} compares TLMs to related work, guided by the rubric of five reasoning principles just discussed. 
% \item 
Sec.~\ref{sec:discussion} concludes with a summary of the contributions of this paper and a discussion of future work.
Due to the space limitations, certain technical details and extensions to the calculi presented in this paper, as well as detailed proofs, are available in the anonymized supplemental material. % Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators. %are also available in the supplement. Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e. \emph{static evaluation}, in  more detail. Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.

@@

\subsection{HTML}
\begin{figure*}[t]
